Acceleration of Global Optimization Algorithm by Detecting Local Extrema Based on Machine Learning

### 2021-09-24 2021/Entropy/entropy.tex
Translate aswers for reviewers

%Russian
Методы машинного обучения активно используются в сочетании с алгоритмами bayesian optimization, основанными на использовании probabilistic surrogate models of the objective function. Детальный обзор данного направления развития методов глобальной оптмизации представлен в \cite{Archetti2019,Zhigljavsky2021}.

@@

Note that the algorithms of the last class do not provide guaranteed convergence to the solution of the problem and are inferior to deterministic algorithms in terms of the quality of solution \cite{Kvasov2018,Sergeyev2018} (e.g., measured by the number of correctly solved problems from a particular set). Therefore, deterministic methods seem to be potentially more effective.

%Russian
Настоящая работа направлена на дальнейшее развитие эффективного детерминированного метода глобальной оптимизации, известного как \textit{the information-statistical global search algorithm} \cite{Strongin2000}. Указанная здесь книга содержит результаты теоретических исследований метода, имеющих непосредственное значение для его практической реализации. В частности, в ней детально рассмотрены вопросы сходимости, выбора параметров и условия остановки алгоритма, и т.д.
Отметим, что алгоритм глобального поиска первоначально был ориентирован на решение unconstrained optimization problems. Впоследствии он был обобщен для решения problems with non-convex constraints\cite{Barkalov2002} and multicriteria optimization problems \cite{Gergel2020}.
Одновременно с этим были предложены различные параллельные варианты указанных алгоритмов, которые могут быть использованы на современных суперкомпьютерах \cite{Barkalov2016,globalizerSystem,Strongin2018}.

@@

%Russian
В целом, можно сказать, что дерево принятия решений -- это бинарное дерево, в листьях которого находятся значения функции, а в остальных узлах -- условия перехода.
В нашем случае при применении дерева решений для построения регрессии каждому узлу соответствует несколько результатов испытаний и значение узла $c_j$, которое вычисляется как среднее квадратичное значений в точках испытаний, приписанных к этому узлу.
Итоговая кусочно-постоянная аппроксимация строится с использованием значений $c_j$, которые находятся в листьях дерева.

Дерево строится рекурсивно, начиная с корневого узла. 
В каждом узле применяется правило принятия решения: все данные  разбиваются на две группы (согласно некоторому правилу разбиения) и отправляются в левый и правый узлы-потомки. Затем процедура рекурсивно разделяет левый и правый узлы. 
Остановка рекурсии в узле происходит в одном из следующих случаев:
\begin{itemize}
	\item Количество приписанных к узлу точе испытаний становится меньше указанного порогового значения (мы использовали 1).
	\item Сумма квадратов отклонений значений функции от значения $c_j$, приписанного к этому узлу, становится меньше заданной точности (мы использовали 0.0001).
\end{itemize}
Подробнее об алгоритме построения регрессии с использованием деревьев решений см., например, \cite{Breiman1984}.

Регрессия, построенная с помощью деревьев решений,
is, on the one hand, rather simple, and on the other hand, it adequately reflects the properties of the function under investigation (the presence or absence of local minima). The selection of regions of attraction of local minima using model (\ref{dtree}) can be organized as follows.

@@

%Russian
Использование additional attribute $q^i, \; 0 \leq i \leq k ,$ позволяет различать точки, полученные во время глобального ($q^i=0$) и во время локального ($q^i=1,2$) поиска. Интервалы, граничные точки которых имеют значение $q^i=1$ или $q^i=2$, будут пропускаться при глобальном поиске, т.е. в них не будут проводиться дальнейшие испытания. 
А выделенное значение $q^i=2$, соответствующее локальному минимуму, будет позволять использовать данные точки при проверке критерия остановки. Если по правилам глобального поиска новая точка испытания должна быть поставлена в заданной $\epsilon$-окрестности одного из найденных локальных минимумов, это будет соответствовать окончанию поиска.

@@

%Russian
Отметим, что указанный критерий остановки проверяется после Step 6, т.е. после вычисления по правилам глобального поиска точки $x^{k+1}$ очередного испытания, но до проведения самого испытания в ней. Смысл данного критерия заключается в следующем. Поиск останавиливается, когда интервал $(x_{t-1},x_t)$, в который попадает точка $x^{k+1}$, становится достаточно малым, либо когда точка $x^{k+1}$ попадает в малую окрестность одного из найденных локальных минимумов, который в этом случае будет одновременно являться и глобальным.

@@

Let us compare the basic global search algorithm (GSA) and its decision tree-based modification (GSA-DT) with the well-known DIRECT global optimization algorithm  \cite{Jones2009}.
%Russian
Выбор именного этого метода для сравнения объясняется следующим. DIRECT является одним из наиболее известных и популярных детерминированных методов решения задач глобальной оптмизации with ``black box'' objective function. Обзор различных модификаций метода, а также примеров решения задач приведен в \cite{Jones2021}.
Для этого метода известно, что при достаточно большом числе поисковых испытаний DIRECT гарантированно найдет глобальное решение задачи.  
Однако если использовать критерий основки по точности, то метод может быстро остановиться в одном из локальных минимумов. Сказанное подтверждается приведенными в данном разделе результатами экспериментов.

Если же говорить про детерминированные оптимизационные алгоритмы, реализованные in popular Computer Algebra Systems (CAS), то их сравнение с такими методами, как DIRECT или GSA, будет некорректным по следующей причине.
Оптимизационные алгоритмы из CAS ориентированы на решение задач с целевой функцией, заданной явно, в виде формулы.
Формульное задание целевой функции предполагает известным также и ее производные, что позволяет использовать методы первого порядка, обладающие значительно более быстрой сходимостью, чем методы нулевого порядка, к которым отноится и DIRECT, и GSA. Но при этом методы первого порядка не гарантируют нахождение глобального минимума. 
Например, система Mathematica предлагает для решения задач глобальной оптмизации методы, которые гарантированно найдут глобальное решение задачи только в случае, если целевая функция и ограничения являются линейными или выпуклыми. Otherwise, the result may sometimes only be a local minimum.

Более того, методы, требующие формульного задания целевой функции, не могут применяться к решению большого класса прикладных задач, в которых вид целевой функции не известен, а ее значения вычисляются в результате численного моделирования. Использование для решения задач подобного вида эвристических методов (таких как Differential Evolution, Simulated Annealing or Random Search) также не всегда приводит к решению задачи. 
These methods often do find the global minimum, but are not guaranteed to do so. В терминах числа корректно решенных задач эвристические методы проигрывают детерминированным \cite{Sergeyev2018}.

@@

Tables \ref{table:average_Shekel} and \ref{table:average_Hill} show the number of search trials that, on average, were required to minimize the Shekel and Hill functions with different search accuracy  $\epsilon$. The number of unsolved problems is indicated in parentheses. 
%Russian
В этих и последующих таблицах для метода GSA-DT приведено полное число испытаний, которые были выполнены как во время глобального, так и во время локального поиска.
При решении задач с точностью $10^{-2}$, $10^{-3}$, $10^{-4}$ число запусков локального поиска при решении одной задачи составляло в среднем 2, 3 и 4 соответственно; а отношение числа испытаний, выполненных по правилам глобального поиска, к числу испытаний, выполненных локальным методом, составляло (при соответствующей точности) 7.2, 6.0, 4.5 для задач класса $F_{HL}$ и 3.2, 2.3, 1.9 для задач класса $F_{SH}$.

@@

%Russian
Еще одним направлением дальнейшей работы будет являться комбинирование предложенного подхода к использованию локального поиска с традиционными способами ускорения алгоритмов глобальной оптмизации. Например, широко известным способом ускорения алгоритмов данного класса является использование local adaptive estimates of the Lipschitz constant $L_i$ в различных зонах $D_i \in D$ области поиска $D$ вместо одной глобальной оценки константы $L$ во всей области $D$ ( see \cite{Kvasov2020,Sergeyev2021}).
Это позволяет алгоритму настраиваться на локальное поведение целевой функции в различных частях допустимой области и сокращает число поисковых испытаний, требующихся для достижения сходимости. Использование только глобальной информации о поведении целевой функции в ходе ее минимизации может значительно замедлить сходимость алгоритма к точке
глобального минимума. В данном направлении также планируется использовать методы машинного обучения для выделения подобластей $D_i \in D$ и построения локальных оценок константы Липшица $L_i$ в выделенных подобластях.

>>>>

%Russian
{ \color{red}
Machine learning methods are actively used in combination with bayesian optimization algorithms based on the probabilistic surrogate models of the objective function. The detailed overview of this trend in the development of global optimization methods is presented in \cite{Archetti2019,Zhigljavsky2021}.
}

@@

Note that the algorithms of the latter class do not provide guaranteed convergence to the solution of the problem and are inferior to deterministic algorithms in terms of the quality of solution \cite{Kvasov2018,Sergeyev2018} (e.g., measured by the number of correctly solved problems from a particular set). Therefore, deterministic methods seem to be potentially more effective.

%Russian
This paper aims to further develop the efficient deterministic global optimization method known as \textit{the information-statistical global search algorithm} \cite{Strongin2000}. The book referenced here contains the results of theoretical studies of the method that are of direct importance for its practical implementation. In particular, it discusses in detail the issues of convergence, the choice of parameters, and the conditions for stopping the algorithm, etc.
Note that the global search algorithm was originally designed for solving unconstrained optimization problems. Later, it was generalized to solve problems with non-convex constraints\cite{Barkalov2002} and multicriteria optimization problems \cite{Gergel2020}.
At the same time, scholars proposed various parallel versions of these algorithms, which can be used on modern supercomputers.\cite{Barkalov2016,globalizerSystem,Strongin2018}.

@@

%Russian
{\color{red}
Generally speaking, a decision tree is a binary tree, the leave nodes of which contain the values of the function, and the other nodes contain the transition conditions.
In our case, when applying a decision tree to construct a regression, each node corresponds to the results of several trials and the value $c_j$, which is calculated as the mean square of the trial points assigned to this node.
The final piecewise constant approximation is constructed using the values $c_j$, located in the leave nodes of the tree.

The tree is built recursively, starting from the root node. 
A decision rule is applied at each node: all data is divided into two groups (according to a partitioning rule) and sent to the left and right child nodes. The procedure then recursively separates the left and right nodes. 
Recursion stops at a node in one of the following cases:
\begin{itemize}
	\item The number of trial points assigned to the node becomes less than the specified threshold value (we used 1).
	\item The sum of the squared deviations of the function values from the value $c_j$, assigned to this node becomes less than the set accuracy (we used $10^{-3}$).
\end{itemize}
For more information on the algorithm for constructing regression using decision trees, see, for example \cite{Breiman1984}.
}

Regression built using decision trees,
is, on the one hand, rather simple, and on the other hand, it adequately reflects the properties of the function under investigation (the presence or absence of local minima). The selection of regions of attraction of local minima using model (\ref{dtree}) can be organized as follows.

@@

%Russian
{\color{red}
Using the additional attribute $q^i, \; 0 \leq i \leq k ,$ makes it possible to distinguish between points obtained during a global ($q^i=0$) and a local ($q^i=1,2$) search. The intervals, the boundary points of which have the value  $q^i=1$ or $q^i=2$, will be skipped in the global search, i.e. they will not be tried further. 
While the selected value $q^i=2$, which corresponds to the local minimum will allow using these points for checking the stopping criterion. When the rules of the global search signal that a new trial point needs be placed in a given  $\epsilon$-neighborhood of one of the found local minima, this will correspond to the end of the search.
}

@@

%Russian
{\color{red}
Note that the specified stopping criterion is checked after Step 6, i.e. after the calculation according to the rules of the global search for the point of the next trial $x^{k+1}$, but before the trial itself is carried out in it. The idea behind this criterion is as follows. The search stops when the interval  $(x_{t-1},x_t)$, in which the point $x^{k+1}$ falls, becomes sufficiently small, or when the point $x^{k+1}$ falls into a small neighborhood of one of the found local minima. Such minima in this case will be simultaneously global.
}

@@

Let us compare the basic global search algorithm (GSA) and its decision tree-based modification (GSA-DT) with the well-known DIRECT global optimization algorithm  \cite{Jones2009}.
{\color{red}
%Russian 
The choice of this particular method for comparison is explained as follows. DIRECT is one of the most well-known and popular deterministic methods for solving global optimization problems with a `` black box '' objective function. An overview of various modifications of the method, as well as examples of solving problems, is given in \cite{Jones2021}.
It is known that with a sufficiently large number of search trials, DIRECT is guaranteed to find a global solution to the problem. 
However, if the stopping by accuracy is used as a criterion, then the method can abruptly stop at one of the local minima, which is confirmed by the experimental results presented in this section.

Regarding the deterministic optimization algorithms implemented in popular Computer Algebra Systems (CAS), their comparison with methods such as DIRECT or GSA will be incorrect for the following reason.
Optimization algorithms from CAS are focused on solving problems with an objective function set explicitly, in the form of a formula.
Formula definition of the objective function assumes that its derivatives are also known, which makes it possible to use first-order methods with significantly faster convergence than zero-order methods, which include both DIRECT and GSA. However, first-order methods do not guarantee finding the global minimum. 
For example, the Mathematica system offers methods for solving global optimization problems that are guaranteed to find a global solution to the problem only if the objective function and constraints are linear or convex. Otherwise, the result may sometimes only be a local minimum.

Moreover, methods that require a formulaic specification of the objective function cannot be applied to the solution of a large class of applied problems in which the form of the objective function is not known, and its values are calculated as a result of numerical simulation. The use of heuristic methods for solving problems of this kind (such as Differential Evolution, Simulated Annealing or Random Search) also does not always lead to the solution of the problem. 
These methods often do find the global minimum, but are not guaranteed to do so. In terms of the number of correctly solved problems, heuristic methods are inferior to deterministic ones \cite{Sergeyev2018}.
}

@@

Tables \ref{table:average_Shekel} and \ref{table:average_Hill} show the number of search trials that, on average, were required to minimize the Shekel and Hill functions with different search accuracy  $\epsilon$. The number of unsolved problems is indicated in parentheses. 
{\color{red}
These and subsequent tables feature the total number of trials that were performed during both global and local searches by GSA-DT.
When solving problems with an accuracy  $10^{-2}$, $10^{-3}$, $10^{-4}$ the number of local search launches when solving one problem on average was equal to 2, 3, and 4, respectively; and the ratio of the number of trials performed according to the global search rules to the number of trials performed by the local method was (with appropriate accuracy) 7.2, 6.0, 4.5 for problems of the $F_{HL}$ class and 3.2, 2.3, 1.9 for the $F_{SH}$ class problems.
}

@@

%Russian
{\color{red}
Another direction for further work will be the combination of the proposed approach to the application of local search with traditional methods of accelerating global optimization algorithms. For example, a well-known way to speed up algorithms of this class is to use local adaptive estimates of the Lipschitz constant $L_i$ in various subdomains $D_i \in D$ within the search domain $D$ instead of a single global estimate of the constant $L$ for the entire domain $D$ ( see \cite{Kvasov2020,Sergeyev2021}).
This allows the algorithm to adjust to the local behavior of the objective function in different parts of the feasible domain, and, thereby, reduces the number of search trials required to achieve convergence. Using only global information about the behavior of the objective function during its minimization can significantly slow down the convergence of the algorithm to a point of global minimum. 
Further research on this issue could focus on using machine learning methods to isolate subdomains $D_i \in D$ and construct local estimates of the Lipschitz constant  $L_i$ in the selected subdomains.
}
